Supplementary Materials for \"COATi: statistical pairwise alignment of protein coding sequences\

### 2024-01-11 alignpair_letter.tex
Update data

>>>>

\subsection*{Empirical Dataset}

Humans and gorillas are two closely related species with very different levels of genome curation. The human reference genome has been revised dozens of times and is currently on version GRCh38.p14, while the gorilla reference genome has only been revised a handful of times and is currently on version gorGor4 \citep[cf.\ ENSEMBL database v110;][]{ensembl_hubbard_2002}. Additionally significant levels of investment have been made to correctly identify and annotate human genes, while gorilla annotations have received limited support in comparison. Together, these reference genomes provide a good opportunity to compare COATi against other aligner as they offer one genome that is high-quality (human) and sister genome that is lower-quality (gorilla).

We used ENSEMBL database v110 \citep{ensembl_hubbard_2002} to create an empirical dataset of protein-coding sequences for both human genes and their gorilla orthologs. We first selected human protein-coding genes that belonged to the Consensus Coding Sequence Set, that were located on an autosomal chromosome, and that had a one-to-one gorilla ortholog. We selected the canonical isoform for both species and removed any pair in which the total nucleotide length was larger larger than 6,000 nucleotides. This resulted in 14,127 sequence pairs and corresponding FASTA files containing CDS sequences. Due to the way that canonical isoforms are identified, there is no guarantee that the isoforms are orthologous even though the genes are. Therefore, a subset of the sequence pairs in this dataset contain human and gorilla sequences with different exon compositions. We have made no attempt to correct these sequence pairs because in our experience genome-wide studies rarely control for such artifacts.
